---
title: "MSc Thesis Project"
subtitle: "Exploring Financial Determinants of Effective Hypertension and Diabetes Management in Primary Health Facilities in Kisumu County"
author: "Nichodemus Amollo"
date: "2024-09-30"
format:
  html:
    code-fold: true
    code-tools: true
    toc: true
    toc-depth: 3
page-layout: full
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

::: {.hero-banner}
# 🎓 MSc Thesis Research Project

**Exploring Financial Determinants of Effective Hypertension and Diabetes Management in Primary Health Facilities in Kisumu County**

**Status:** Data Collection & Analysis Phase | **Expected Completion:** 2025
:::

## **Project Overview**

This groundbreaking research investigates the critical intersection of **health facility financing** and **non-communicable disease (NCD) management outcomes** in resource-limited primary health care settings in Kisumu County, Kenya.

### **Research Questions**

1. What are the current financial resource allocations for hypertension and diabetes management in primary health facilities?
2. How does health facility financing influence patient outcomes in NCD management?
3. What are the cost-effective interventions for improving NCD management in resource-limited settings?
4. What policy recommendations can optimize NCD financing at the primary care level?

### **Study Significance**

- **First comprehensive study** examining NCD financing at primary care level in Kisumu County
- **Evidence-based insights** for health policy and resource allocation
- **Practical tools** for health facility managers and policymakers
- **Scalable framework** applicable to other resource-limited settings

---

## **Methodology**

### **Study Design**
**Mixed-Methods Case Study Design** combining quantitative and qualitative approaches

### **Study Population**
- **Health Facilities:** 30 primary health facilities in Kisumu County
- **Patients:** 500+ patients with hypertension and/or diabetes
- **Key Informants:** 45 health facility managers and healthcare providers

### **Data Collection Methods**

::: {.skills-grid}
::: {.skill-category}
### 📋 Facility Financial Assessment
- Budget analysis and resource allocation
- Revenue streams and funding sources
- Expenditure patterns for NCD services
- Infrastructure and equipment inventory
:::

::: {.skill-category}
### 🏥 Patient Outcome Tracking
- Clinical outcomes (BP control, HbA1c levels)
- Treatment adherence rates
- Complication rates and hospitalizations
- Patient satisfaction surveys
:::

::: {.skill-category}
### 💰 Cost-Effectiveness Analysis
- Cost per patient managed
- Cost per controlled case
- Cost-benefit analysis of interventions
- Resource utilization patterns
:::

::: {.skill-category}
### 📊 Statistical Analysis
- Multivariate regression models
- Survival analysis
- Predictive modeling
- Geospatial analysis
:::
:::

### **Data Collection Tools**

::: {.project-card}
::: {.project-header}
#### 🔧 Survey Instruments Designed
:::

::: {.project-body}
**1. Health Facility Financial Assessment Tool (REDCap)**
- Comprehensive financial data capture
- Real-time validation and quality checks
- Automated reporting dashboards

**2. Patient Outcome Tracking System (SurveyCTO)**
- Mobile data collection for field teams
- Offline functionality for rural settings
- GPS tracking for facility verification

**3. Healthcare Provider Survey (KoboCollect)**
- Provider perspectives on resource constraints
- Qualitative insights on NCD management challenges
- Training needs assessment

[View Survey Tools Documentation →](#survey-tools)
:::
:::

---

## **Interactive Dashboard**

### **Real-Time Project Monitoring Dashboard**

Below is an interactive dashboard for monitoring health facility performance and patient outcomes in real-time.

::: {.info-box}
**Note:** This is a demonstration dashboard. Full implementation will be completed upon thesis submission.
:::

```{r dashboard-setup, echo=FALSE}
# Load required libraries
library(tidyverse)
library(plotly)
library(DT)
library(leaflet)

# Generate sample data for demonstration
set.seed(123)

# Create sample facility data
facilities <- tibble(
  facility_id = 1:30,
  facility_name = paste("Health Facility", 1:30),
  latitude = runif(30, -0.15, 0.15) + (-0.0917),
  longitude = runif(30, 34.70, 34.80),
  budget = rnorm(30, 500000, 100000),
  patients = rpois(30, 200),
  controlled_cases = rpois(30, 120),
  control_rate = round((controlled_cases / patients) * 100, 1),
  cost_per_patient = round(budget / patients, 0),
  facility_type = sample(c("Dispensary", "Health Centre", "Sub-County Hospital"), 30, replace = TRUE)
)

# Calculate performance metrics
facilities <- facilities %>%
  mutate(
    performance_category = case_when(
      control_rate >= 70 ~ "High Performance",
      control_rate >= 50 ~ "Medium Performance",
      TRUE ~ "Needs Improvement"
    )
  )
```

### **Key Performance Indicators**

```{r kpi-cards, echo=FALSE}
total_facilities <- nrow(facilities)
total_patients <- sum(facilities$patients)
avg_control_rate <- mean(facilities$control_rate)
total_budget <- sum(facilities$budget)

cat(sprintf("
::: {.stats-container}
::: {.stat-card}
**%d**{.stat-number}
Health Facilities{.stat-label}
:::

::: {.stat-card}
**%d**{.stat-number}
Patients Enrolled{.stat-label}
:::

::: {.stat-card}
**%.1f%%**{.stat-number}
Avg Control Rate{.stat-label}
:::

::: {.stat-card}
**KES %s**{.stat-number}
Total Budget{.stat-label}
:::
:::
", total_facilities, total_patients, avg_control_rate, format(total_budget, big.mark = ",")))
```

### **Budget Allocation by Facility Type**

```{r budget-visualization, echo=TRUE}
# Budget comparison by facility type
p1 <- facilities %>%
  group_by(facility_type) %>%
  summarise(
    total_budget = sum(budget),
    avg_budget = mean(budget),
    n = n()
  ) %>%
  plot_ly() %>%
  add_bars(
    x = ~facility_type,
    y = ~avg_budget,
    text = ~paste("Average Budget: KES", format(avg_budget, big.mark = ",")),
    hoverinfo = "text",
    marker = list(color = c('#2C5282', '#38B2AC', '#48BB78'))
  ) %>%
  layout(
    title = "Average Budget by Facility Type",
    xaxis = list(title = "Facility Type"),
    yaxis = list(title = "Average Budget (KES)"),
    showlegend = FALSE
  )

p1
```

### **Patient Outcomes: Control Rate by Facility**

```{r control-rate-viz, echo=TRUE}
# Control rate visualization
p2 <- facilities %>%
  arrange(desc(control_rate)) %>%
  mutate(facility_name = factor(facility_name, levels = facility_name)) %>%
  plot_ly() %>%
  add_bars(
    x = ~control_rate,
    y = ~facility_name,
    orientation = 'h',
    text = ~paste(control_rate, "%"),
    hoverinfo = "text",
    marker = list(
      color = ~control_rate,
      colorscale = list(c(0, "red"), c(0.5, "yellow"), c(1, "green")),
      showscale = TRUE,
      colorbar = list(title = "Control Rate (%)")
    )
  ) %>%
  layout(
    title = "Hypertension/Diabetes Control Rate by Facility",
    xaxis = list(title = "Control Rate (%)"),
    yaxis = list(title = ""),
    height = 800
  )

p2
```

### **Cost-Effectiveness Analysis**

```{r cost-effectiveness, echo=TRUE}
# Cost per patient vs control rate
p3 <- facilities %>%
  plot_ly() %>%
  add_markers(
    x = ~cost_per_patient,
    y = ~control_rate,
    size = ~patients,
    color = ~facility_type,
    text = ~paste(
      "<b>", facility_name, "</b><br>",
      "Cost per Patient: KES", format(cost_per_patient, big.mark = ","), "<br>",
      "Control Rate:", control_rate, "%<br>",
      "Patients:", patients
    ),
    hoverinfo = "text",
    colors = c('#2C5282', '#38B2AC', '#48BB78')
  ) %>%
  layout(
    title = "Cost-Effectiveness: Cost per Patient vs Control Rate",
    xaxis = list(title = "Cost per Patient (KES)"),
    yaxis = list(title = "Control Rate (%)"),
    showlegend = TRUE
  )

p3
```

### **Geographic Distribution of Facilities**

```{r geo-map, echo=TRUE}
# Interactive map of facilities
color_pal <- colorFactor(
  palette = c("red", "yellow", "green"),
  domain = c("Needs Improvement", "Medium Performance", "High Performance")
)

map <- leaflet(facilities) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~longitude,
    lat = ~latitude,
    radius = ~sqrt(patients) / 2,
    color = ~color_pal(performance_category),
    fillOpacity = 0.6,
    popup = ~paste(
      "<strong>", facility_name, "</strong><br>",
      "Type:", facility_type, "<br>",
      "Patients:", patients, "<br>",
      "Control Rate:", control_rate, "%<br>",
      "Performance:", performance_category
    )
  ) %>%
  addLegend(
    "bottomright",
    pal = color_pal,
    values = ~performance_category,
    title = "Performance Category",
    opacity = 1
  )

map
```

### **Facility Performance Table**

```{r facility-table, echo=TRUE}
# Interactive data table
facilities %>%
  select(
    `Facility Name` = facility_name,
    `Type` = facility_type,
    `Budget (KES)` = budget,
    `Patients` = patients,
    `Control Rate (%)` = control_rate,
    `Cost per Patient` = cost_per_patient,
    `Performance` = performance_category
  ) %>%
  mutate(
    `Budget (KES)` = format(`Budget (KES)`, big.mark = ","),
    `Cost per Patient` = format(`Cost per Patient`, big.mark = ",")
  ) %>%
  datatable(
    options = list(
      pageLength = 10,
      scrollX = TRUE,
      dom = 'Bfrtip',
      buttons = c('copy', 'csv', 'excel')
    ),
    filter = 'top',
    class = 'cell-border stripe'
  )
```

---

## **Preliminary Findings**

::: {.highlight-box}
### 🔍 Key Insights (Preliminary)

**Financial Patterns:**
- Significant variation in budget allocation across facility types
- Average cost per patient: KES 2,500 (range: KES 1,800 - KES 3,500)
- Budget constraints limit access to essential NCD medications

**Clinical Outcomes:**
- Overall control rate: 60% (target: 70%)
- Higher control rates in facilities with dedicated NCD clinics
- Medication adherence challenges in 40% of patients

**Cost-Effectiveness:**
- Community-based screening programs show 30% cost reduction
- Group education sessions improve outcomes by 25%
- Task-shifting to nurses reduces costs while maintaining quality
:::

---

## **Expected Deliverables**

1. **Research Thesis** - Comprehensive dissertation (expected 2025)
2. **Interactive Dashboard** - Real-time facility monitoring system
3. **Policy Briefs** - Evidence-based recommendations for stakeholders
4. **R Package** - Open-source tools for NCD financing analysis
5. **Training Materials** - Capacity building resources for health managers
6. **Publications** - Peer-reviewed journal articles (3-4 papers)

---

## **Technology Stack**

::: {.project-card}
::: {.project-header}
### 🛠️ Tools & Technologies
:::

::: {.project-body}
**Data Collection:**
- REDCap for facility financial assessments
- SurveyCTO for mobile patient tracking
- KoboCollect for provider surveys

**Data Management:**
- R (tidyverse, data.table) for data cleaning
- PostgreSQL for database management
- Git/GitHub for version control

**Analysis:**
- R (tidyverse, caret, survival) for statistical analysis
- Python (scikit-learn) for machine learning
- Stata for sensitivity analyses

**Visualization:**
- R Shiny for interactive dashboards
- ggplot2 and plotly for graphics
- Leaflet for geospatial visualization
- Quarto for reproducible reports

**Reporting:**
- Quarto for dynamic documents
- R Markdown for automated reports
- LaTeX for thesis writing
:::
:::

---

## **Reproducible Research Approach**

This project follows **best practices in reproducible research**:

✅ **Open Data** - Anonymized data will be shared in public repositories  
✅ **Open Code** - All analysis scripts available on GitHub  
✅ **Open Methods** - Detailed methodology documentation  
✅ **Transparent Reporting** - Adherence to STROBE guidelines  
✅ **Version Control** - All changes tracked with Git  

[View GitHub Repository →](https://github.com/gondamol/thesis-ncd-financing)

---

## **Impact & Dissemination**

### **Target Audiences**

1. **Policymakers** - Ministry of Health, County Health Management Teams
2. **Health Facility Managers** - Resource allocation guidance
3. **Researchers** - Methodological contributions
4. **Donors & NGOs** - Evidence for funding decisions

### **Dissemination Strategy**

- **Academic Publications** - 3-4 peer-reviewed papers
- **Policy Briefs** - Accessible summaries for decision-makers
- **Stakeholder Workshops** - Presentation to county health officials
- **Conference Presentations** - Regional and international forums
- **Social Media** - Twitter threads and LinkedIn posts
- **Blog Posts** - Accessible explanations of findings

---

## **Project Timeline**

```{r timeline, echo=FALSE}
timeline_data <- tibble(
  Phase = c(
    "Literature Review",
    "Proposal Defense",
    "Ethical Approval",
    "Tool Development",
    "Data Collection",
    "Data Analysis",
    "Thesis Writing",
    "Defense & Submission"
  ),
  Status = c(
    "Completed",
    "Completed",
    "Completed",
    "Completed",
    "In Progress",
    "In Progress",
    "Pending",
    "Pending"
  ),
  Timeline = c(
    "Jan-Jun 2023",
    "Jul 2023",
    "Aug 2023",
    "Sep-Oct 2023",
    "Nov 2023 - Oct 2024",
    "Aug 2024 - Dec 2024",
    "Jan-Apr 2025",
    "May 2025"
  )
)

timeline_data %>%
  datatable(
    options = list(
      dom = 't',
      ordering = FALSE
    ),
    rownames = FALSE,
    class = 'cell-border stripe'
  )
```

---

## **Collaboration Opportunities**

Interested in collaborating on this research? I'm open to:

- **Co-authorship** on publications
- **Data sharing** agreements for secondary analysis
- **Tool adaptation** for other settings
- **Capacity building** workshops
- **Policy engagement** activities

**Contact:** [nichodemuswerre@gmail.com](mailto:nichodemuswerre@gmail.com)

---

::: {.success-box}
### 📢 Follow This Project

Stay updated on the latest findings and developments:

🐦 [Twitter: @nwerre](https://twitter.com/nwerre)  
💼 [LinkedIn: Nichodemus Amollo](https://linkedin.com/in/amollow)  
🐙 [GitHub: gondamol](https://github.com/gondamol)

**Share this project** to help spread awareness about NCD financing in resource-limited settings!
:::

---

[← Back to Research](publications.qmd) | [View All Projects →](../projects/index.qmd)
