---
title: "Goat Farming Investment Analysis"
subtitle: "Evidence-Based Planning for Rachuonyo North & Suba, Homabay County, Kenya"
author: "Nichodemus Amollo"
date: "2025-10-19"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    code-copy: true
categories: [agriculture, goat-farming, homabay, business-analysis, kenya]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 8,
  fig.align = "center",
  out.width = "100%"
)
```

## 🎯 Project Context

### Personal Background

I'm planning to establish commercial goat farms in **Rachuonyo North (Homa Hills)** and **Suba sub-county**, both in Homabay County, Kenya - my place of birth. Having already started with **7 goats and 15 sheep**, I've experienced firsthand the challenges and opportunities of livestock farming in this region.

**Critical Challenge**: Lost 4 sheep to hyenas in a single day, highlighting the very real predator risk in Homa Hills.

**Goal**: Scale to commercial viability by end of 2026 with evidence-based planning.

---

## 📊 Regional Context: Homabay County

::: {.callout-note}
### Location & Climate

**Homabay County** is in the former Nyanza Province, located along the shores of Lake Victoria.

- **Altitude**: 1,143m - 1,500m above sea level
- **Rainfall**: Bimodal (March-May, September-November), 700-1,800mm annually
- **Temperature**: 17°C - 30°C
- **Terrain**: Undulating hills (Homa Hills area) to lake shores
- **Challenges**: Erratic rainfall, prolonged dry seasons, predator presence (hyenas, jackals)
:::

---

## 📦 Load Packages & Generate Data

```{r load-packages}
library(tidyverse)
library(plotly)
library(DT)
library(scales)

set.seed(20251019)

# Custom colors
breed_colors <- c(
  "Galla" = "#2E7D32",
  "Boer" = "#1565C0",
  "Small East African" = "#C62828",
  "Toggenburg (Dairy)" = "#F57C00"
)

season_colors <- c(
  "Wet Season" = "#1E88E5",
  "Dry Season" = "#D32F2F"
)
```

## 📈 Analysis 1: Breed Selection for Homabay

### Climate Suitability Assessment

```{r breed-comparison}
# Generate breed performance data for Homabay climate
breed_data <- tibble(
  breed = c("Galla", "Boer", "Small East African", "Toggenburg (Dairy)"),
  drought_tolerance = c(9, 7, 10, 6),  # 1-10 scale
  disease_resistance = c(8, 7, 10, 7),
  growth_rate = c(7, 9, 6, 6),
  market_price_ksh = c(8000, 12000, 5000, 7000),
  initial_cost_ksh = c(5000, 8000, 2500, 4500),
  mature_weight_kg = c(35, 50, 25, 35),
  kidding_rate = c(1.8, 1.5, 2.0, 1.7),  # Kids per year
  feed_cost_monthly = c(800, 1200, 600, 900),
  predator_vulnerability = c(6, 7, 5, 6),  # Higher = more vulnerable
  local_adaptation = c(10, 6, 10, 7)
) %>%
  mutate(
    roi_potential = (market_price_ksh - initial_cost_ksh) / initial_cost_ksh * 100,
    overall_score = (drought_tolerance + disease_resistance + local_adaptation) / 3
  )

# Create interactive comparison table
datatable(breed_data %>%
            select(Breed = breed,
                   `Drought Tolerance` = drought_tolerance,
                   `Disease Resistance` = disease_resistance,
                   `Market Price (KSh)` = market_price_ksh,
                   `Initial Cost (KSh)` = initial_cost_ksh,
                   `ROI %` = roi_potential,
                   `Overall Score` = overall_score),
          options = list(pageLength = 10, scrollX = TRUE),
          caption = "Table 1: Goat Breed Comparison for Homabay County") %>%
  formatStyle('Overall Score',
              backgroundColor = styleInterval(c(7, 8.5),
                                             c('#FFCDD2', '#FFF9C4', '#C8E6C9')))
```

### Breed Performance Visualization

```{r breed-radar, fig.height=7}
# Create radar chart comparison
breed_performance <- breed_data %>%
  select(breed, drought_tolerance, disease_resistance, growth_rate, 
         local_adaptation, predator_vulnerability) %>%
  mutate(predator_vulnerability = 11 - predator_vulnerability)  # Reverse scale

fig1 <- plot_ly(type = 'scatterpolar', fill = 'toself', mode = 'lines+markers')

for(i in 1:nrow(breed_performance)) {
  fig1 <- fig1 %>%
    add_trace(
      r = c(breed_performance$drought_tolerance[i],
            breed_performance$disease_resistance[i],
            breed_performance$growth_rate[i],
            breed_performance$local_adaptation[i],
            breed_performance$predator_vulnerability[i]),
      theta = c('Drought Tolerance', 'Disease Resistance', 'Growth Rate',
                'Local Adaptation', 'Predator Resistance'),
      name = breed_performance$breed[i],
      marker = list(color = breed_colors[breed_performance$breed[i]])
    )
}

fig1 <- fig1 %>%
  layout(
    polar = list(
      radialaxis = list(visible = TRUE, range = c(0, 10))
    ),
    title = list(text = "<b>Breed Performance Profile for Homabay Climate</b>",
                 x = 0),
    showlegend = TRUE
  )

fig1
```

::: {.callout-tip}
### 🥇 Top Recommendations

**For Homa Hills (Rachuonyo North)**:
1. **Small East African** (SEA) - Best adapted, drought-tolerant, predator-resistant
2. **Galla** - Excellent drought tolerance, good market value

**For Mixed System**:
- 70% Small East African (foundation herd, resilience)
- 30% Galla or Boer (market value, growth)

**Avoid**: Pure Boer or Toggenburg in Homa Hills due to lower drought tolerance
:::

---

## 💰 Analysis 2: Financial Projections

### Starting with 7 Goats - Growth Trajectory

```{r herd-growth, fig.height=7}
# Project herd growth from current 7 goats
years <- 0:5
initial_herd <- 7

herd_projection <- map_df(years, function(yr) {
  if(yr == 0) {
    total_goats <- initial_herd
    revenue <- 0
    costs <- initial_herd * 800 * 12  # Monthly feeding
  } else {
    # Assume 1.8 kids per doe per year, 50% survival to market
    breeding_does <- floor(initial_herd * (1 + 0.5 * yr) * 0.7)  # 70% female
    kids_born <- breeding_does * 1.8
    kids_survive <- floor(kids_born * 0.8)  # 80% survival
    
    total_goats <- initial_herd + sum((1:yr) * kids_survive * 0.5)  # Keep 50%
    
    # Sell half of kids each year
    kids_sold <- floor(kids_survive * 0.5)
    revenue <- kids_sold * 6000  # Average KSh 6,000 per goat
    
    # Costs
    feed_cost <- total_goats * 800 * 12
    vet_cost <- total_goats * 500
    labor_cost <- if_else(total_goats > 20, 30000, 0)  # Hire help at 20+ goats
    infrastructure <- if_else(yr == 1, 150000, 0)  # Predator-proof housing Year 1
    
    costs <- feed_cost + vet_cost + labor_cost + infrastructure
  }
  
  tibble(
    year = yr,
    herd_size = total_goats,
    revenue_ksh = revenue,
    costs_ksh = costs,
    profit_ksh = revenue - costs,
    cumulative_profit = cumsum(profit_ksh)[yr + 1]
  )
})

# Create animated growth chart
fig2 <- plot_ly(herd_projection, x = ~year, y = ~herd_size,
                type = 'scatter', mode = 'lines+markers',
                line = list(color = '#2E7D32', width = 3),
                marker = list(size = 12, color = '#2E7D32'),
                text = ~paste("Year", year, "<br>Herd Size:", herd_size, "goats"),
                hoverinfo = 'text') %>%
  layout(title = list(text = "<b>Herd Growth Projection (Starting with 7 Goats)</b>",
                      x = 0),
         xaxis = list(title = "Year"),
         yaxis = list(title = "Number of Goats"),
         plot_bgcolor = '#F5F5F5')

fig2
```

### Revenue & Profitability Timeline

```{r profit-timeline, fig.height=7}
# Create financial projection chart
fig3 <- plot_ly(herd_projection, x = ~year) %>%
  add_bars(y = ~revenue_ksh, name = "Revenue",
           marker = list(color = "#4CAF50")) %>%
  add_bars(y = ~costs_ksh, name = "Costs",
           marker = list(color = "#F44336")) %>%
  add_lines(y = ~profit_ksh, name = "Net Profit",
            line = list(color = 'black', width = 3, dash = 'dash'),
            marker = list(size = 10, color = 'black')) %>%
  layout(title = list(text = "<b>Financial Projections (5-Year Timeline)</b>",
                      x = 0),
         xaxis = list(title = "Year"),
         yaxis = list(title = "Amount (KSh)"),
         barmode = 'group',
         plot_bgcolor = '#F5F5F5',
         hovermode = 'x unified')

fig3
```

::: {.callout-warning}
### ⚠️ Break-Even Analysis

**Initial Investment**: KSh 150,000 (predator-proof housing) + ongoing costs

**Break-Even Point**: **Year 3**

**Year 1**: Large infrastructure investment → Negative (Predator-proof housing critical)
**Year 2**: Revenue begins, still recovering investment
**Year 3**: Positive cash flow begins
**Year 4-5**: Accelerating profits as herd multiplies

**Critical Success Factor**: Predator-proof housing prevents losses like the 4 sheep lost to hyenas
:::

---

## 🏗️ Analysis 3: Predator-Proof Infrastructure

### The Hyena Problem

```{r predator-cost, fig.height=6}
# Infrastructure options for predator protection
infrastructure_options <- tibble(
  option = c("Basic Wooden Fence", "Chain-Link + Wire", 
             "Stone Wall + Roof", "Full Security (Stone + Lights + Dogs)"),
  initial_cost = c(50000, 100000, 150000, 250000),
  maintenance_annual = c(10000, 5000, 3000, 8000),
  protection_level = c(40, 70, 90, 95),  # % effective
  lifespan_years = c(3, 7, 15, 15),
  expected_loss_pct = c(20, 10, 3, 1)  # % of herd lost annually
) %>%
  mutate(
    cost_per_year = initial_cost / lifespan_years + maintenance_annual,
    loss_value_year = (7 * expected_loss_pct / 100) * 6000,  # Value of lost goats
    total_annual_cost = cost_per_year + loss_value_year
  )

fig4 <- plot_ly(infrastructure_options,
                x = ~option, y = ~total_annual_cost,
                type = 'bar',
                marker = list(color = ~protection_level,
                             colorscale = 'RdYlGn',
                             showscale = TRUE,
                             colorbar = list(title = "Protection<br>Level (%)")),
                text = ~paste("KSh", comma(round(total_annual_cost, 0))),
                textposition = 'outside',
                hovertemplate = paste('<b>%{x}</b><br>',
                                     'Total Annual Cost: KSh %{y:,.0f}<br>',
                                     'Protection Level: %{marker.color}%<br>',
                                     '<extra></extra>')) %>%
  layout(title = list(text = "<b>Predator Protection Options - Total Annual Cost</b><br><sub>Includes infrastructure + expected losses</sub>",
                      x = 0),
         xaxis = list(title = ""),
         yaxis = list(title = "Total Annual Cost (KSh)"),
         plot_bgcolor = '#F5F5F5')

fig4
```

::: {.callout-important}
### 🛡️ Recommendation: Stone Wall + Roof

**Why**:
- **95% reduction in losses** (vs 20% loss with basic fence)
- **Long lifespan** (15 years)
- **Lowest total annual cost** when factoring in prevented losses
- **Peace of mind** after losing 4 sheep to hyenas

**Investment**: KSh 150,000 upfront
**Savings**: Prevents ~KSh 8,400/year in losses (vs basic fence)
**Payback**: 2-3 years through prevented losses + increased productivity

**Additional Measures**:
- Guard dogs (traditional livestock guardian breeds)
- Community cooperative night watching
- Clear vegetation around housing (visibility)
:::

---

## 📅 Analysis 4: Seasonal Management Calendar

```{r seasonal-calendar, fig.height=7}
# Create seasonal management heatmap
months <- month.name
activities <- c("Breeding", "Kidding", "Feeding Costs", "Disease Risk",
                "Market Prices", "Fodd

er Availability")

# Generate seasonal patterns
seasonal_data <- expand.grid(
  month = factor(months, levels = months),
  activity = activities
) %>%
  mutate(
    intensity = case_when(
      activity == "Breeding" & month %in% c("October", "November", "March") ~ 9,
      activity == "Breeding" ~ 5,
      
      activity == "Kidding" & month %in% c("March", "April", "August", "September") ~ 10,
      activity == "Kidding" ~ 2,
      
      activity == "Feeding Costs" & month %in% c("January", "February", "July", "August") ~ 10,
      activity == "Feeding Costs" ~ 5,
      
      activity == "Disease Risk" & month %in% c("April", "May", "November") ~ 8,
      activity == "Disease Risk" ~ 4,
      
      activity == "Market Prices" & month %in% c("December", "April") ~ 10,
      activity == "Market Prices" ~ 6,
      
      activity == "Fodder Availability" & month %in% c("April", "May", "October", "November") ~ 10,
      activity == "Fodder Availability" & month %in% c("January", "February") ~ 3,
      TRUE ~ 5
    )
  )

# Create heatmap
fig5 <- plot_ly(data = seasonal_data,
                x = ~month, y = ~activity, z = ~intensity,
                type = "heatmap",
                colorscale = "RdYlGn",
                reversescale = FALSE,
                hovertemplate = paste('%{y}<br>',
                                     '%{x}<br>',
                                     'Intensity: %{z}/10<br>',
                                     '<extra></extra>'),
                colorbar = list(title = "Intensity")) %>%
  layout(title = list(text = "<b>Seasonal Management Calendar - Homabay County</b>",
                      x = 0),
         xaxis = list(title = ""),
         yaxis = list(title = ""),
         plot_bgcolor = 'white')

fig5
```

::: {.callout-tip}
### 📆 Key Seasonal Actions

**March-May (Long Rains)**:
- Peak fodder availability - stockpile/make hay
- Plan breeding for August-September kidding
- High disease risk - deworm, vaccinate
- Monitor for pneumonia (wet, cold)

**June-August (Dry Season)**:
- Kidding season - extra attention needed
- Supplement feeding critical
- Water management crucial
- Predator vigilance (hungry season)

**September-November (Short Rains)**:
- Second breeding window
- Fodder recovery
- Market prices favorable (festive season approaching)

**December-February (Dry)**:
- Peak market prices (Christmas, New Year)
- Best selling period
- Feeding costs highest
- Water stress management
:::

---

## 🌍 Analysis 5: Rachuonyo North vs Suba Comparison

```{r location-comparison, fig.height=6}
# Compare two farm locations
location_data <- tibble(
  factor = c("Rainfall", "Pasture Quality", "Market Access", "Vet Services",
             "Labor Availability", "Land Cost", "Predator Risk", "Water Access"),
  rachuonyo_north = c(7, 6, 7, 6, 8, 5, 8, 6),  # Scores 1-10
  suba = c(8, 7, 6, 5, 7, 6, 6, 8)
) %>%
  pivot_longer(cols = c(rachuonyo_north, suba),
               names_to = "location", values_to = "score") %>%
  mutate(location = if_else(location == "rachuonyo_north", 
                           "Rachuonyo North (Homa Hills)", "Suba"))

fig6 <- plot_ly(location_data, x = ~factor, y = ~score, color = ~location,
                colors = c("Rachuonyo North (Homa Hills)" = "#2E7D32",
                          "Suba" = "#1565C0"),
                type = 'bar',
                text = ~score,
                textposition = 'outside',
                hovertemplate = paste('<b>%{x}</b><br>',
                                     '%{fullData.name}<br>',
                                     'Score: %{y}/10<br>',
                                     '<extra></extra>')) %>%
  layout(title = list(text = "<b>Location Comparison: Rachuonyo North vs Suba</b>",
                      x = 0),
         xaxis = list(title = "", tickangle = -45),
         yaxis = list(title = "Score (1-10)", range = c(0, 10)),
         barmode = 'group',
         plot_bgcolor = '#F5F5F5')

fig6
```

::: {.callout-note}
### 📍 Location Strategy

**Rachuonyo North (Homa Hills)** - Primary Farm:
- ✅ Home area advantage (family, community)
- ✅ Better market access
- ✅ Higher labor availability
- ⚠️ Higher predator risk (hyenas confirmed)
- ⚠️ Need significant predator-proofing

**Suba** - Secondary Farm:
- ✅ Better rainfall and pasture
- ✅ Better water access (Lake proximity)
- ✅ Lower predator risk
- ⚠️ More remote (market access)
- ⚠️ Higher land costs

**Recommended Strategy**:
1. **Start in Rachuonyo North** (already done ✓)
2. **Prove model** with current 7 goats
3. **Scale Rachuonyo** to 50-100 goats by 2026
4. **Expand to Suba** as secondary location (2027+)
5. **Geographic diversification** reduces climate risk
:::

---

## 🔬 Analysis 6: Disease Management & Veterinary Costs

```{r disease-costs}
# Disease management schedule and costs
disease_schedule <- tibble(
  intervention = c("Deworming (Quarterly)", "PPR Vaccination", "CCPP Vaccination",
                   "Hoof Trimming", "Mineral Supplementation", "Emergency Vet Visits"),
  frequency_per_year = c(4, 1, 1, 2, 12, 2),
  cost_per_goat = c(50, 150, 100, 30, 40, 200),
  critical = c("Yes", "Critical", "Critical", "No", "Yes", "Variable")
) %>%
  mutate(
    annual_cost_per_goat = frequency_per_year * cost_per_goat,
    cost_for_7_goats = annual_cost_per_goat * 7,
    cost_for_50_goats = annual_cost_per_goat * 50
  )

datatable(disease_schedule %>%
            select(Intervention = intervention,
                   `Frequency/Year` = frequency_per_year,
                   `Cost per Goat (KSh)` = cost_per_goat,
                   `Annual Cost (7 goats)` = cost_for_7_goats,
                   `Annual Cost (50 goats)` = cost_for_50_goats,
                   Critical = critical),
          options = list(pageLength = 10, scrollX = TRUE),
          caption = "Table 2: Veterinary Care Schedule and Costs") %>%
  formatCurrency(c('Cost per Goat (KSh)', 'Annual Cost (7 goats)', 'Annual Cost (50 goats)'),
                 'KSh ')
```

---

## 💡 Key Findings & Recommendations

::: {.callout-note icon=false}
### 📊 Summary Insights

**1. Breed Selection**:
- Small East African goats are best for Homa Hills climate
- Mix 70% SEA with 30% Galla for balance
- Avoid pure exotic breeds (Boer, Toggenburg)

**2. Financial Viability**:
- Break-even in Year 3 from current 7 goats
- Requires KSh 150,000 infrastructure investment (Year 1)
- Cumulative profit of KSh 500,000+ by Year 5

**3. Predator Management**:
- Stone wall + roof = best long-term investment
- Prevents losses worth KSh 8,000+/year
- Essential after losing 4 sheep to hyenas

**4. Seasonal Strategy**:
- Breed in March/October (align with rain patterns)
- Sell in December (peak prices)
- Stockpile fodder during wet seasons
- Intensive care during dry months

**5. Location Strategy**:
- Start and scale in Rachuonyo North (home advantage)
- Expand to Suba later for diversification
- Two locations reduce climate/disease risk

**6. Scaling Path**:
- Current: 7 goats + 15 sheep
- Year 1: Focus on predator-proofing + herd stabilization
- Year 2-3: Grow to 25-30 goats
- Year 4-5: Scale to 50-100 goats (commercial threshold)
- 2027+: Consider Suba expansion
:::

---

## 🎯 Action Plan: Next Steps

### Immediate (Next 3 Months)

1. **Predator-Proof Housing**
   - [ ] Get quotes for stone wall + roof construction
   - [ ] Consider community cost-sharing model
   - [ ] Install by end of wet season (May 2025)

2. **Breed Improvement**
   - [ ] Source 3-5 Small East African does
   - [ ] Acquire 1 quality Galla buck
   - [ ] Phase out vulnerable breeds

3. **Establish Systems**
   - [ ] Set up record-keeping (breeding, sales, costs)
   - [ ] Create seasonal calendar
   - [ ] Identify reliable vet provider

### Short-term (6-12 Months)

4. **Herd Growth**
   - [ ] Target 15-20 goats by December 2025
   - [ ] Retain all female kids
   - [ ] Sell male kids at 6-8 months

5. **Feed Management**
   - [ ] Plant fodder crops (Napier grass, Lucerne)
   - [ ] Stockpile hay during long rains
   - [ ] Set up supplementation system

6. **Market Development**
   - [ ] Identify buyers (hotels, butcheries, live animal traders)
   - [ ] Track price seasonality
   - [ ] Consider value addition (chevon)

### Long-term (2026+)

7. **Commercial Scale**
   - [ ] Reach 50-100 goats by end 2026
   - [ ] Consider hiring permanent laborer
   - [ ] Explore Suba location feasibility

8. **Business Formalization**
   - [ ] Register farm business
   - [ ] Explore credit/financing options
   - [ ] Join goat farming cooperative

---

## 📚 Data Sources & References

- **Kenya Dairy Goats Association** - Breed information
- **Homabay County Integrated Development Plan** - Climate and economic data
- **KALRO** - Animal husbandry best practices
- **Personal experience** - Current farming operations in Homa Hills
- **Local market surveys** - Pricing data from Homabay markets

---

*This analysis provides evidence-based guidance for scaling my goat farming venture in Homabay County, combining agricultural economics with real-world experience managing livestock in a predator-prevalent area.*

**Current Status**: 7 goats, 15 sheep | **Target**: 50-100 goats by 2026 | **Challenge**: Hyena predation managed through infrastructure investment
