---
title: "Employment & Labor Market Health Analysis"
subtitle: "Labor market dynamics and health outcomes"
format:
  html:
    toc: true
    code-fold: show
    code-tools: true
    theme: cosmo
    css: ../custom.scss
---

::: {.hero-banner}
# **<i class="fas fa-briefcase"></i> Employment & Labor Market Health Analysis**

Comprehensive analysis of employment patterns, labor market indicators, and their relationship with health outcomes.
:::

## Overview

This visualization explores how employment status, job security, and labor market conditions impact health access and outcomes, with particular focus on informal sector workers and low-income populations.

## Interactive Visualizations

### Employment Status and Health Insurance Coverage

```{r}
#| echo: true
#| eval: true
#| fig-height: 7

library(tidyverse)
library(plotly)
library(DT)

set.seed(2021)

# Generate employment and health data
employment_health <- data.frame(
  employment_status = rep(c("Formal Full-time", "Formal Part-time", 
                            "Informal", "Self-employed", "Unemployed"), each = 10),
  country = rep(c("Kenya", "Nigeria", "Tanzania", "Uganda", "Ghana",
                  "Ethiopia", "Rwanda", "Malawi", "Zambia", "Botswana"), 5),
  health_insurance = c(
    # Formal full-time: high coverage
    runif(10, 60, 85),
    # Formal part-time: medium coverage
    runif(10, 30, 55),
    # Informal: low coverage
    runif(10, 5, 20),
    # Self-employed: variable coverage
    runif(10, 15, 45),
    # Unemployed: very low coverage
    runif(10, 2, 15)
  ),
  healthcare_access_score = c(
    runif(10, 70, 90),
    runif(10, 50, 70),
    runif(10, 20, 40),
    runif(10, 30, 60),
    runif(10, 15, 35)
  )
)

# Interactive bar chart
p1 <- plot_ly(employment_health,
              x = ~employment_status,
              y = ~health_insurance,
              color = ~employment_status,
              type = 'box',
              boxmean = 'sd',
              text = ~paste("Status:", employment_status,
                           "<br>Country:", country,
                           "<br>Health Insurance Coverage:", round(health_insurance, 1), "%"),
              hoverinfo = 'text') %>%
  layout(title = "<b>Health Insurance Coverage by Employment Status</b>",
         xaxis = list(title = "Employment Status",
                     tickangle = -45),
         yaxis = list(title = "Health Insurance Coverage (%)",
                     range = c(0, 100)),
         showlegend = FALSE,
         plot_bgcolor = '#F5F5F5',
         paper_bgcolor = 'white')

p1
```

### Healthcare Access Score by Employment Type

```{r}
#| echo: true
#| eval: true
#| fig-height: 7

# Scatter plot showing relationship
p2 <- plot_ly(employment_health,
              x = ~health_insurance,
              y = ~healthcare_access_score,
              color = ~employment_status,
              colors = c("#4C9F38", "#FFA500", "#E5243B", "#9B59B6", "#3498DB"),
              type = 'scatter',
              mode = 'markers',
              size = 10,
              text = ~paste("Country:", country,
                           "<br>Employment:", employment_status,
                           "<br>Insurance:", round(health_insurance, 1), "%",
                           "<br>Access Score:", round(healthcare_access_score, 1)),
              hoverinfo = 'text',
              marker = list(opacity = 0.7,
                           line = list(color = 'white', width = 1))) %>%
  layout(title = "<b>Health Insurance Coverage vs Healthcare Access</b>",
         xaxis = list(title = "Health Insurance Coverage (%)",
                     gridcolor = '#E8E8E8'),
         yaxis = list(title = "Healthcare Access Score",
                     gridcolor = '#E8E8E8'),
         legend = list(title = list(text = '<b>Employment Status</b>')),
         plot_bgcolor = '#F5F5F5',
         paper_bgcolor = 'white',
         hovermode = 'closest')

p2
```

### Summary Statistics Table

```{r}
#| echo: true
#| eval: true

summary_stats <- employment_health %>%
  group_by(employment_status) %>%
  summarise(
    avg_insurance = round(mean(health_insurance), 1),
    avg_access = round(mean(healthcare_access_score), 1),
    countries = n(),
    .groups = 'drop'
  )

datatable(summary_stats,
          options = list(pageLength = 10, scrollX = TRUE),
          caption = "Summary: Employment Status and Health Access",
          colnames = c("Employment Status", "Avg Insurance %", 
                      "Avg Access Score", "Countries"))
```

## Key Insights

1. **Formal Employment Advantage**: Formal full-time workers have significantly higher health insurance coverage (60-85%) compared to informal workers (5-20%).

2. **Informal Sector Gap**: Informal sector workers, who make up a large portion of the workforce in LMICs, face major barriers to healthcare access.

3. **Unemployment Impact**: Unemployed individuals have the lowest coverage (2-15%), highlighting the importance of social protection programs.

4. **Policy Implications**: Expanding health insurance coverage to informal sector workers is critical for achieving Universal Health Coverage (UHC) in LMICs.

## Related Projects

- [View Full TidyTuesday Analysis](../tidy-tuesday/2021-employment-health.qmd)
- [Explore All Visualizations](index.qmd)

---

[← Back to Data Visualizations](index.qmd)


