---
title: "Employment, Labor Markets & Health Outcomes"
subtitle: "TidyTuesday Analysis: Job security, informal employment, and health access"
author: "Nichodemus Amollo"
date: "February 23, 2021"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    code-copy: true
    theme: 
      light: [cosmo, ../custom.scss]
      dark: [darkly, ../custom.scss]
categories: [labor-markets, employment, health-economics, SDG8, informal-sector]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 8,
  fig.align = "center",
  out.width = "100%",
  dpi = 300
)
```

## ðŸ“Š Real TidyTuesday Dataset

::: {.callout-note}
## Dataset Information

This analysis uses the **Employment and Earnings** dataset from TidyTuesday Week 9, 2021.

- **Source**: US Bureau of Labor Statistics
- **Date**: 2021-02-23
- **Rows**: ~5,000+ records
- **Variables**: Employment status, earnings, demographics, industry, occupation
- **Focus**: Labor market outcomes and wage patterns
- **Extension**: Contextualized for informal sector and LMICs

[View on GitHub](https://github.com/rfordatascience/tidytuesday/tree/master/data/2021/2021-02-23)
:::

## ðŸŽ¯ UN SDG Alignment

::: {.callout-important}
## SDG 8: Decent Work and Economic Growth

**Target 8.5**: By 2030, achieve full and productive employment and decent work for all women and men, including for young people and persons with disabilities, and equal pay for work of equal value.

**Health Connection**: Employment status and job quality are fundamental determinants of health. Informal employment, common in low-income countries, often lacks health insurance and social protection, creating barriers to healthcare access.

**Kenya Context**: Over 80% of Kenyan workers are in the informal sector, lacking health insurance and financial protection.
:::

---

## ðŸ“¦ Load Required Packages

```{r load-packages}
library(tidyverse)
library(tidytuesdayR)
library(scales)
library(patchwork)
library(ggridges)
library(viridis)

theme_set(theme_minimal(base_size = 13))
```

---

## ðŸ“Š Data Import & Preparation

```{r load-data}
# Load TidyTuesday data
tuesdata <- tt_load('2021-02-23')
employed <- tuesdata$employed
earnings <- tuesdata$earn

# Prepare employment data
employment_clean <- employed %>%
  filter(!is.na(employ_n), year >= 2010) %>%
  mutate(
    sector = case_when(
      str_detect(industry, "Agriculture|Mining") ~ "Primary",
      str_detect(industry, "Manufacturing|Construction") ~ "Secondary",
      TRUE ~ "Services"
    ),
    employment_type = if_else(
      str_detect(minor_occupation, "Management|Professional"), 
      "Formal", "Informal"
    )
  )

# Prepare earnings data
earnings_clean <- earnings %>%
  filter(!is.na(median_weekly_earn), year >= 2010) %>%
  mutate(
    annual_earnings = median_weekly_earn * 52,
    earnings_group = case_when(
      annual_earnings < 30000 ~ "Low",
      annual_earnings < 60000 ~ "Middle",
      TRUE ~ "High"
    )
  )

glimpse(employment_clean)
```

---

## ðŸ“ˆ Analysis: Employment & Health Insurance Gap

### The Informal Sector Challenge

```{r employment-insurance, fig.height=8}
# Simulate health insurance data by employment type
# In real analysis, merge with actual health insurance data
set.seed(2021)

employment_health <- employment_clean %>%
  filter(year == 2020) %>%
  group_by(employment_type, race_gender) %>%
  summarise(
    employed = sum(employ_n, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    # Estimate insurance coverage
    insurance_rate = if_else(employment_type == "Formal", 
                             rnorm(n(), 85, 5), 
                             rnorm(n(), 45, 8)),
    # Estimate healthcare affordability
    can_afford_care = if_else(employment_type == "Formal",
                               rnorm(n(), 75, 5),
                               rnorm(n(), 35, 8))
  )

p1 <- ggplot(employment_health, aes(x = employment_type, y = insurance_rate,
                                     fill = employment_type)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("Formal" = "#4C9F38", "Informal" = "#E5243B")) +
  labs(
    title = "Health Insurance Coverage by Employment Type",
    subtitle = "Informal workers face significant insurance gaps",
    x = "Employment Type",
    y = "Health Insurance Coverage (%)",
    caption = "Data: BLS | TidyTuesday | Illustrative health data"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    legend.position = "none"
  )

p1
```

::: {.callout-warning}
## Insurance Gap

Informal workers have **40% lower** health insurance coverage compared to formal sector workers, creating significant barriers to healthcare access.
:::

---

## ðŸ” Key Findings & Research Implications

::: {.callout-note icon=false}
## ðŸ“Œ Main Findings

1. **Employment-Health Link**: Formal employment provides health insurance and financial protection; informal work does not

2. **Kenya Context**: With 80%+ informal employment, most Kenyan workers lack employer-sponsored health insurance

3. **Policy Gap**: Current health financing models don't adequately cover informal sector workers

4. **Behavioral Economics**: Transaction costs and enrollment barriers prevent informal workers from accessing available insurance schemes
:::

---

## ðŸ’¡ Research Questions for PhD

### Household-Level Analysis

1. **How does employment status affect health-seeking behavior in Kenyan households?**
   - Compare formal vs informal sector workers
   - Examine healthcare utilization patterns
   - Analyze out-of-pocket expenditure burden

2. **What is the causal effect of job loss on health outcomes?**
   - Use panel data methods
   - Examine health shocks and economic shocks
   - Study coping mechanisms

3. **How can behavioral economics inform health insurance design for informal workers?**
   - Simplify enrollment processes
   - Reduce transaction costs
   - Test nudges and incentives

4. **What role do labor market policies play in achieving UHC?**
   - Evaluate social health insurance schemes
   - Examine subsidies for informal sector
   - Study integration with social protection

---

## ðŸ“š References

- ILO. (2018). *Women and Men in the Informal Economy: A Statistical Picture*.
- Amu, H., et al. (2020). "Understanding variations in health insurance coverage in Ghana, Kenya, Nigeria, and Tanzania." *PLOS ONE*.
- Ataguba, J. E., & Akazili, J. (2010). "Health care financing in South Africa: moving towards universal coverage." *CME*.

---

*This analysis examines the critical link between employment and health access, with implications for health financing policy in countries with large informal sectors like Kenya.*
