---
title: "Global Temperature Anomalies"
subtitle: "TidyTuesday Analysis: Historical temperature changes and climate trends"
author: "Nichodemus Amollo"
date: "January 2, 2024"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    code-copy: true
    theme: 
      light: [cosmo, ../custom.scss]
      dark: [darkly, ../custom.scss]
    css: ../styles.scss
categories: [climate, SDG13, temperature, global-warming]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 8,
  fig.align = "center",
  out.width = "100%",
  dpi = 300
)
```

## 📊 Real TidyTuesday Dataset

::: {.callout-note}
## Dataset Information

This analysis uses the **Global Surface Temperatures** dataset from TidyTuesday Week 28, 2023.

- **Source**: NASA GISS Surface Temperature Analysis
- **Date**: 2023-07-11
- **Rows**: ~2,000+ records
- **Variables**: Year, month, temperature anomaly (°C), region
- **Focus**: Long-term global temperature trends and climate change

[View on GitHub](https://github.com/rfordatascience/tidytuesday/tree/master/data/2023/2023-07-11)
:::

## 🎯 UN SDG Alignment

::: {.callout-important}
## SDG 13: Climate Action

**Target 13.1**: Strengthen resilience and adaptive capacity to climate-related hazards.

**Paris Agreement**: Limit global warming to well below 2°C, preferably 1.5°C above pre-industrial levels.

This analysis examines historical temperature anomalies to understand the pace and pattern of global warming.
:::

---

# Global Temperature Changes

This analysis explores global temperature anomalies over time, examining trends in global warming and their implications for climate change.

## 📦 Load Required Packages

```{r load-packages}
library(tidyverse)      # Data manipulation and visualization
library(scales)         # Scale functions
library(lubridate)      # Date handling
library(viridis)        # Color palettes

# Set theme
theme_set(theme_minimal(base_size = 13))
```

---

## Data Preparation

```{r load-data}
# Load real TidyTuesday data
library(tidytuesdayR)

# Method 1: Using tidytuesdayR package (recommended)
tuesdata <- tt_load('2023-07-11')
temperature_data <- tuesdata$global_temps

# Method 2: Direct URL (if package unavailable)
# temperature_data <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-07-11/global_temps.csv')

# Quick exploration
glimpse(temperature_data)

# Clean and prepare data
temp_clean <- temperature_data %>%
  # Filter for annual averages
  filter(!is.na(annual_anomaly)) %>%
  # Create decade variable
  mutate(
    decade = floor(year / 10) * 10,
    era = case_when(
      year < 1950 ~ "Pre-1950",
      year < 1980 ~ "1950-1979",
      year < 2000 ~ "1980-1999",
      TRUE ~ "2000+"
    )
  )

# Preview the data
head(temp_clean)
```



```{r}
# Example plot
ggplot(temperature_data, aes(x = year, y = temperature_change)) +
  geom_line() +
  labs(title = "Global Temperature Changes Over Time", x = "Year", y = "Temperature Change (°C)")
```


