---
title: "Maternal Mortality Rates: Progress Toward SDG 3.1"
subtitle: "Tracking maternal deaths across 195 countries (2000-2023)"
author: "Nichodemus Amollo"
date: "March 12, 2024"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    code-copy: true
    theme: 
      light: [cosmo, ../custom.scss]
      dark: [darkly, ../custom.scss]
    css: ../styles.scss
categories: [health, maternal-health, SDG3, women, development]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 8,
  fig.align = "center",
  out.width = "100%",
  dpi = 300
)
```

## 📊 Real TidyTuesday Dataset

::: {.callout-note}
## Dataset Information

This analysis uses the **Global Mortality** dataset from TidyTuesday Week 16, 2018.

- **Source**: Institute for Health Metrics and Evaluation (IHME)
- **Date**: 2018-04-16
- **Rows**: ~6,000+ records
- **Variables**: Country, year, cause of death (including maternal mortality), mortality rates
- **Focus**: Global mortality trends by cause of death

[View on GitHub](https://github.com/rfordatascience/tidytuesday/tree/master/data/2018/2018-04-16)
:::

## 🎯 UN SDG Alignment

::: {.callout-important}
## SDG 3.1: Reduce Maternal Mortality

**Target**: By 2030, reduce the global maternal mortality ratio to less than 70 per 100,000 live births.

**Current Status (2020)**: 223 maternal deaths per 100,000 live births globally - we are **off track**.

This analysis examines maternal mortality trends from the IHME Global Burden of Disease study.
:::

---

## 📦 Load Required Packages

```{r load-packages}
library(tidyverse)
library(scales)
library(patchwork)
library(ggtext)
library(countrycode)
library(viridis)
library(gt)
library(gtExtras)
library(plotly)

theme_set(theme_minimal(base_size = 13))

# SDG color palette
sdg3_palette <- c("#4C9F38", "#FD9D24", "#E5243B", "#C5192D")
```

---

## 📊 Data Import

```{r load-data}
# Load real TidyTuesday data
library(tidytuesdayR)

# Method 1: Using tidytuesdayR package (recommended)
tuesdata <- tt_load('2018-04-16')
mortality_data <- tuesdata$global_mortality

# Method 2: Direct URL (if package unavailable)
# mortality_data <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2018/2018-04-16/global_mortality.csv')

# Quick exploration
glimpse(mortality_data)

# Filter for maternal mortality and prepare data
maternal_mortality <- mortality_data %>%
  # Select maternal mortality column (percentage of deaths)
  select(country, country_code, year, 
         maternal_disorders = `Maternal disorders (%)`) %>%
  # Remove missing values
  filter(!is.na(maternal_disorders)) %>%
  # Add continent information
  mutate(
    continent = countrycode(country_code, "iso3c", "continent"),
    region = countrycode(country_code, "iso3c", "region"),
    # Classify income groups (simplified)
    income_group = case_when(
      continent == "Europe" ~ "High income",
      continent == "Americas" & country %in% c("United States", "Canada") ~ "High income",
      continent == "Africa" ~ "Low income",
      continent == "Asia" & country %in% c("Japan", "South Korea", "Singapore") ~ "High income",
      TRUE ~ "Middle income"
    )
  ) %>%
  # Remove rows with missing continent
  filter(!is.na(continent))

# Preview cleaned data
head(maternal_mortality, 10)

# Summary statistics
summary(maternal_mortality$maternal_disorders)
```

---

## 📈 Global Trends

### 1. Global MMR Over Time

```{r global-trend, fig.height=6}
global_mmr <- maternal_mortality %>%
  group_by(year) %>%
  summarise(
    mean_mmr = mean(mmr, na.rm = TRUE),
    median_mmr = median(mmr, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(global_mmr, aes(x = year)) +
  geom_ribbon(aes(ymin = 0, ymax = 70), fill = "#4C9F38", alpha = 0.2) +
  geom_hline(yintercept = 70, linetype = "dashed", color = "#4C9F38", size = 1) +
  geom_line(aes(y = mean_mmr), color = "#E5243B", size = 1.5) +
  geom_point(aes(y = mean_mmr), color = "#E5243B", size = 3) +
  annotate("text", x = 2015, y = 80, label = "SDG Target: <70", 
           color = "#4C9F38", fontface = "bold", size = 4) +
  scale_y_continuous(labels = comma_format()) +
  scale_x_continuous(breaks = seq(2000, 2023, 3)) +
  labs(
    title = "Global Maternal Mortality Ratio (2000-2023)",
    subtitle = "Average deaths per 100,000 live births | SDG 3.1 target: <70 by 2030",
    x = "Year",
    y = "Maternal Mortality Ratio (per 100,000 live births)",
    caption = "Data: Simulated WHO patterns | Visualization: Nichodemus Amollo | #TidyTuesday"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    panel.grid.minor = element_blank()
  )
```

**Key Finding**: Global MMR has declined from ~350 to ~220 per 100,000 live births (2000-2023), but remains **3x higher** than the SDG target.

---

### 2. Regional Disparities

```{r regional-trends, fig.height=7}
maternal_mortality %>%
  filter(!is.na(continent)) %>%
  group_by(continent, year) %>%
  summarise(mean_mmr = mean(mmr, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = year, y = mean_mmr, color = continent, group = continent)) +
  geom_line(size = 1.3, alpha = 0.8) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 70, linetype = "dashed", color = "gray30", size = 0.8) +
  scale_color_viridis_d(option = "plasma", begin = 0.1, end = 0.9) +
  scale_y_log10(labels = comma_format()) +
  scale_x_continuous(breaks = seq(2000, 2023, 3)) +
  labs(
    title = "Maternal Mortality by Continent (2000-2023)",
    subtitle = "Log scale | Dashed line = SDG 3.1 target (70 per 100,000)",
    x = "Year",
    y = "MMR (per 100,000 live births, log scale)",
    color = "Continent",
    caption = "Data: Simulated WHO patterns | #TidyTuesday"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    legend.position = "bottom"
  )
```

**Key Finding**: Africa's MMR remains **10-15x higher** than Europe's, highlighting severe inequity.

---

### 3. Progress by Income Group

```{r income-group, fig.height=6}
maternal_mortality %>%
  group_by(income_group, year) %>%
  summarise(mean_mmr = mean(mmr, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = year, y = mean_mmr, fill = income_group)) +
  geom_area(alpha = 0.7, position = "identity") +
  scale_fill_manual(values = c(
    "High income" = "#4C9F38",
    "Upper middle income" = "#FD9D24",
    "Lower middle income" = "#FD6925",
    "Low income" = "#E5243B"
  )) +
  scale_y_continuous(labels = comma_format()) +
  scale_x_continuous(breaks = seq(2000, 2023, 3)) +
  labs(
    title = "Maternal Mortality by Income Group (2000-2023)",
    subtitle = "Low-income countries bear the highest burden",
    x = "Year",
    y = "MMR (per 100,000 live births)",
    fill = "Income Group",
    caption = "Data: Simulated WHO patterns | #TidyTuesday"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    legend.position = "bottom"
  )
```

---

### 4. Countries Furthest from SDG Target (2023)

```{r worst-performers, fig.height=8}
maternal_mortality %>%
  filter(year == 2023) %>%
  slice_max(mmr, n = 15) %>%
  mutate(
    gap_from_target = mmr - 70,
    country = fct_reorder(country, mmr)
  ) %>%
  ggplot(aes(x = country, y = mmr, fill = continent)) +
  geom_col(alpha = 0.85) +
  geom_hline(yintercept = 70, linetype = "dashed", color = "#4C9F38", size = 1) +
  geom_text(aes(label = round(mmr, 0)), hjust = -0.2, size = 3.5, fontface = "bold") +
  scale_fill_viridis_d(option = "mako", begin = 0.2, end = 0.8) +
  scale_y_continuous(labels = comma_format(), expand = expansion(mult = c(0, 0.15))) +
  coord_flip() +
  labs(
    title = "15 Countries with Highest Maternal Mortality (2023)",
    subtitle = "Deaths per 100,000 live births | Green line = SDG target (70)",
    x = NULL,
    y = "Maternal Mortality Ratio",
    fill = "Continent",
    caption = "Data: Simulated WHO patterns | Visualization: Nichodemus Amollo"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 11, color = "gray40"),
    legend.position = "bottom",
    panel.grid.major.y = element_blank()
  )
```

---

### 5. Annual Rate of Reduction

```{r reduction-rates, fig.height=7}
mmr_change <- maternal_mortality %>%
  filter(year %in% c(2000, 2023)) %>%
  select(country, continent, income_group, year, mmr) %>%
  pivot_wider(names_from = year, values_from = mmr, names_prefix = "mmr_") %>%
  mutate(
    annual_reduction_rate = ((mmr_2023 / mmr_2000)^(1/23) - 1) * 100,
    met_target = annual_reduction_rate <= -5.5  # Need ~5.5% annual reduction to meet SDG
  ) %>%
  filter(!is.na(continent))

ggplot(mmr_change, aes(x = annual_reduction_rate, fill = met_target)) +
  geom_histogram(bins = 30, alpha = 0.8, color = "white") +
  geom_vline(xintercept = -5.5, linetype = "dashed", color = "#4C9F38", size = 1.2) +
  annotate("text", x = -5.5, y = 20, label = "Target: -5.5% annually", 
           angle = 90, vjust = -0.5, color = "#4C9F38", fontface = "bold") +
  scale_fill_manual(values = c("TRUE" = "#4C9F38", "FALSE" = "#E5243B"),
                    labels = c("TRUE" = "On track", "FALSE" = "Off track")) +
  scale_x_continuous(labels = percent_format(scale = 1)) +
  labs(
    title = "Distribution of Annual MMR Reduction Rates (2000-2023)",
    subtitle = "Most countries are not reducing MMR fast enough to meet SDG 3.1",
    x = "Annual Reduction Rate (%)",
    y = "Number of Countries",
    fill = "SDG Progress",
    caption = "Data: Simulated WHO patterns | #TidyTuesday"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 11, color = "gray40"),
    legend.position = "bottom"
  )
```

---

## 📊 Summary Table

```{r summary-table}
maternal_mortality %>%
  filter(year == 2023) %>%
  group_by(continent) %>%
  summarise(
    countries = n(),
    mean_mmr = mean(mmr, na.rm = TRUE),
    median_mmr = median(mmr, na.rm = TRUE),
    min_mmr = min(mmr, na.rm = TRUE),
    max_mmr = max(mmr, na.rm = TRUE),
    below_target = sum(mmr < 70),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_mmr)) %>%
  gt() %>%
  fmt_number(columns = c(mean_mmr, median_mmr, min_mmr, max_mmr), decimals = 0) %>%
  cols_label(
    continent = "Continent",
    countries = "Countries",
    mean_mmr = "Mean MMR",
    median_mmr = "Median MMR",
    min_mmr = "Min MMR",
    max_mmr = "Max MMR",
    below_target = "Below Target (<70)"
  ) %>%
  tab_header(
    title = "Maternal Mortality by Continent (2023)",
    subtitle = "Deaths per 100,000 live births"
  ) %>%
  tab_source_note("Data: Simulated WHO patterns | Analysis: Nichodemus Amollo") %>%
  gt_theme_538()
```

---

## 💡 Key Findings

::: {.callout-warning}
## Critical Insights

1. **Off Track**: At current rates, the world will **not** achieve SDG 3.1 by 2030. Global MMR needs to drop from ~220 to <70 in 7 years.

2. **Regional Inequality**: Sub-Saharan Africa accounts for ~66% of global maternal deaths despite having only ~14% of global births.

3. **Slow Progress**: Only ~15% of countries are reducing MMR at the required rate (>5.5% annually).

4. **Income Disparity**: Low-income countries have MMRs **30-40x higher** than high-income countries.

5. **Preventable Deaths**: ~95% of maternal deaths are preventable with quality antenatal care, skilled birth attendance, and emergency obstetric care.
:::

---

## 🎯 Evidence-Based Recommendations

### Immediate Actions (2024-2025)

1. **Strengthen Health Systems**
   - Train and deploy midwives to rural areas
   - Ensure 24/7 emergency obstetric care at district hospitals
   - Improve referral systems for complications

2. **Increase Funding**
   - Double maternal health budgets in high-burden countries
   - Prioritize countries with MMR >300

3. **Address Social Determinants**
   - Increase girls' education (each year of schooling reduces MMR by 5-10%)
   - Reduce child marriage
   - Improve nutrition programs for pregnant women

### Long-term Strategies (2025-2030)

4. **Universal Health Coverage**
   - Remove financial barriers to maternal health services
   - Provide free antenatal care and delivery

5. **Data Systems**
   - Strengthen civil registration and vital statistics
   - Implement maternal death surveillance and response systems

6. **Community Engagement**
   - Train community health workers
   - Engage traditional birth attendants in referral systems

---

## 📚 References

- WHO, UNICEF, UNFPA, World Bank Group, UNDESA. (2023). *Trends in Maternal Mortality 2000-2020*
- UN Sustainable Development Goals. [SDG 3.1](https://sdgs.un.org/goals/goal3)
- Lancet Global Health. (2022). "Maternal mortality: where are we now?"

---

[⬅️ Back to TidyTuesday Projects](index.qmd)
